---
title: "Carbon reactivity index - modelling"
output: html_notebook
---

# Install packages

```{r packages, message=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here,
       terra,
       dplyr,
       caret,
       ggplot2,
       sf,
       CAST,
       lwgeom,
       geosphere,
       quantregForest,
       doParallel,
       ModelMetrics,
       DescTools,
       forcats)
```


# Preparation

## Load required data

```{r load_data}
predictors <- rast(here("1_predictors", "data", "output", "env_vars_0.01deg.tif"))
predictors <- predictors[[1:11]] # Remove winter SPM

#The one created in 4_CRI_model Aa_data_prep 
CRI <- vect(here("4_CRI_model", "data", "interim", "CRI.shp"))

rm_resp <- read.csv(here("4_CRI_model", "data", "interim", "rm_resp.csv"))
summary(rm_resp)
```


## Which response variable are we predicting?

Define response variable, unit and data:

core: OCdata_ukShelf.csv
ll: unprojected (latitude and longitude)
0.01: resolution of predictors in degrees

```{r response_type}
resp_var <- "CRI"
resp_unit <- "-"
dat <- "core_ll_0.01"
```


# Quantile Regression Forest model

## Distances in environmental space

Distances in environmental (feature) space are computed. This measures how different two locations are in terms of predictor variables
Is a multivariate dissimilarity analysis and is unitless. What it does is to compare each sample environment (surrounsings) differs from 
the rest of the predictor grid, or locations we want to predict.

```{r env_space_dist, message=FALSE}
dist_env <- geodist(st_as_sf(CRI), 
                    predictors,
                    type = "feature"
                     )

plot(dist_env)
plot(dist_env) + scale_x_log10()
```


## Distances in geographic space

Distances in geographic space are computed. This measures the physical separation on the Earth in terms of m or km
between the sample locations and the points in the model domain (locations we want to predict).

```{r geogr_space_dist, message=FALSE}
dist_geogr <- geodist(st_as_sf(CRI), 
                     predictors,
                     type = "geo"
                     )

plot(dist_geogr)
plot(dist_geogr, unit="km") + scale_x_log10()
```


## Creating spatial blocks

Spatial blocks and folds are created. The folds will be used in a spatial k-fold cross validation. The k-fold nearest neighbour distance matching algorithm is used here.

```{r knndm}
k <- 5 # Number of folds
knndmfolds <- knndm(tpoints = st_as_sf(CRI),
                    modeldomain = predictors,
                    k = k,
                    samplesize = 2000)
```


## Distances in geographic space including CV distances

Provides the relevant distances for each test fold to train data. This allow us to assess
how spatially separated each test fold is from the prediction domain avoiding the risk of overfitting.

```{r geogr_space_dist2, message=FALSE}
dist_geogr2 <- geodist(st_as_sf(CRI), 
                       predictors,
                       cvfolds= knndmfolds$indx_test,
                       type = "geo"
                       )

plot(dist_geogr2)
plot(dist_geogr2, unit="km") + scale_x_log10()
```


## Model tuning

A Quantile Regression Forest model is tuned. Predictor variables are selected in a forward feature selection approach and various values of the mtry parameter are tested in a spatial k-fold cross validation.

The maximum number of iterations to be performed can be calculated upfront, based on the number of pre-selected predictors:

```{r max_iter}
factorial(length(names(predictors)))/(factorial(2)*factorial(length(names(predictors))-2)) + sum(c((length(names(predictors))-2):1))
```


### Forward feature selection

The best combination of predictor variables (features) is found in a forward feature selection process.

```{r ffs, message=FALSE, warning=FALSE}
nCores <- detectCores()
cl <- makePSOCKcluster(nCores - 1)
registerDoParallel(cl)

set.seed(42)

model <- ffs(rm_resp[names(predictors)],
               rm_resp$CRIndex,
               metric = "Rsquared",
               method="qrf",
               what = 0.5,
               replace = FALSE,
               importance = TRUE,
               trControl = trainControl(method="CV",
                                        number = k,
                                        savePredictions = "final",
                                        index = knndmfolds$indx_train, 
                                        allowParallel = TRUE),
               verbose = TRUE)

stopCluster(cl)

model

sel_preds <- model$selectedvars
```


### FFS plot

Plot of R2 over the model runs.

```{r ffs_plot}
plot(model)
```


## Validation statistics

The validation results of the optimal RF model.

Note that these are the statistics based on the predicted values of the selected model. These differ from the values from the tuning (above), which are the means of the k predictions based on the folds.

```{r validation_stats}
t <- data.frame(model$pred$pred, model$pred$obs)

ccc <- CCC(x = t$model.pred.obs, y = t$model.pred.pred) # Concordance Correlation Coefficient

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(sum(t$model.pred.obs - t$model.pred.pred)/nrow(t), 3)
validation[1,2] <- round(rmse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,3] <- round(cor(t$model.pred.obs, t$model.pred.pred)^2, 3)
validation[1,4] <- round(ccc$rho.c$est, 3)

colnames(validation) <- c("ME", "RMSE", "r2", "CCC")
rownames(validation) <- NULL
validation
```
validation
      ME  RMSE    r2   CCC
1 -0.005 0.071 0.158 0.305

## Histograms of observed and predicted organic carbon content (weight-%)

```{r histogram_oc, message=FALSE}
hist <- ggplot(t) + 
  geom_histogram(aes(x = model.pred.obs), bins = 30, fill = "#69b3a2", alpha = 0.5, position = "identity") +
  geom_histogram(aes(x = model.pred.pred), bins = 30, fill = "#404080", alpha = 0.5, position = "identity") +
  annotate("text", x = 0.95, y = 500, label = "observed", colour = "#69b3a2") +
  annotate("text", x = 0.95, y = 470, label = "predicted", colour = "#404080") +
  theme_bw() +
  xlab(paste0(resp_var, " (", resp_unit, ")")) +
  ylab("count")
  
hist

jpeg(filename = here("4_CRI_model", "figures", paste0(resp_var, "_obs_vs_pred_histogram.jpg")), width = 15, height = 10, units = "cm", res = 300)
hist
dev.off()
```


## Validation plot

```{r validation_plot, message=FALSE}
p <-  ggplot(t, aes(x = model.pred.pred, y = model.pred.obs)) +
      geom_hex(bins = 50) +
      #geom_point() +
      scale_fill_continuous(type = "viridis") +
      geom_abline(intercept = 0, slope = 1, colour = "grey", linewidth = 1) +
      geom_smooth(method = "lm") +
      theme_bw() +
      scale_x_continuous(name = "Predicted value") +
      scale_y_continuous(name = "Observed value") +
      ggtitle(paste0(resp_var, " (", resp_unit, ")"))

p

jpeg(filename = here("4_CRI_model", "figures", paste0("validation_plot_",resp_var, ".jpg")), width = 15, height = 10, units = "cm", res = 300)
p
dev.off()
```


## Variable importance

```{r variable_importance_plot, warning=FALSE}
imp <- varImp(model$finalModel, scale = FALSE)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL

importance_col <- names(imp)[1]
imp <- imp[order(imp[[importance_col]], decreasing = TRUE), c("Predictor", importance_col)]
colnames(imp)[2] <- "IncMSE"

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    theme_bw()
    
impfig

jpeg(filename = here("4_CRI_model", "figures", paste0("variable_importance_plot_",resp_var, ".jpg")), width = 10, height = 10, units = "cm", res = 300)
impfig
dev.off()
```


## Distances in environmental space including CV distances

```{r env_space_dist2, message=FALSE}
dist_env2 <- geodist(st_as_sf(CRI), 
                     predictors,
                     type = "feature",
                     cvfolds= knndmfolds$indx_test,
                     variables = sel_preds
                     )

plot(dist_env2)
plot(dist_env2) + scale_x_log10()
```


## Partial dependence

Partial dependence plots give a graphical depiction of the marginal effect of a variable on the response.

```{r partial_plots}
m2 <- model$finalModel
class(m2) <- "randomForest"

for (i in 1:length(sel_preds)) {
  partialPlot(x = m2, pred.data = rm_resp, x.var = sel_preds[i], main = "", xlab = sel_preds[i], ylab = paste0(resp_var, " (", resp_unit, ")"))
}

```


# Predict QRF model

## Predict CRI

CRI is predicted. Median values of the QRF distribution are calculated as central values. The 90% prediction interval and the prediction interval ratio are calculated as measures of uncertainty.

```{r predict_oc}
preds <- raster::stack(predictors[[sel_preds]])
CRI_med <- predict(preds, model$finalModel, what = 0.5)
CRI_p5 <- predict(preds, model$finalModel, what = 0.05)
CRI_p25 <- predict(preds, model$finalModel, what = 0.25)
CRI_p75 <- predict(preds, model$finalModel, what = 0.75)
CRI_p95 <- predict(preds, model$finalModel, what = 0.95)

```


## Area of applicability

```{r aoa}
CRI_trainDI <- trainDI(model = model,
                        variables = sel_preds)
print(CRI_trainDI)

CRI_aoa <- aoa(newdata = predictors, 
                model = model,
                trainDI = CRI_trainDI,
                variables = sel_preds
               )

plot(CRI_aoa)

fr <- freq(CRI_aoa$AOA)
print(paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels"))
```


## Plot results

```{r plot_results}
plot(rast(CRI_p5), main = "CRI P5")
plot(rast(CRI_p25), main = "CRI P25")
plot(rast(CRI_med), main = "CRI median")
plot(rast(CRI_p75), main = "CRI P75")
plot(rast(CRI_p95), main = "CRI P95")
plot(CRI_aoa$DI, main = "Dissimilarity index")
plot(CRI_aoa$AOA, main = "Area of applicability")
```


## Convert AOA from raster to polygon

```{r aoa_poly}
aoa_poly <- as.polygons(CRI_aoa$AOA, dissolve = TRUE)
plot(aoa_poly)

write_sf(st_as_sf(aoa_poly), dsn = here("4_CRI_model", "data", "output"), layer = paste0(resp_var, "_", dat, "_AOA_", Sys.Date()), driver = "ESRI Shapefile")
```


## Export results

```{r export_results}
writeRaster(CRI_med, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_median_", Sys.Date(), ".tif"))
writeRaster(CRI_p5, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_P5_", Sys.Date(), ".tif"))
writeRaster(CRI_p25, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_P25_", Sys.Date(), ".tif"))
writeRaster(CRI_p75, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_P75_", Sys.Date(), ".tif"))
writeRaster(CRI_p95, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_P95_", Sys.Date(), ".tif"))
writeRaster(CRI_aoa$DI, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_DI_", Sys.Date(), ".tif"))
writeRaster(CRI_aoa$AOA, paste0(here("4_CRI_model", "data", "output"), "/", resp_var, "_", dat, "_AOA_", Sys.Date(), ".tif"))
```


## Output a log file

```{r log}
sink(file = paste0(here("4_CRI_model", "data", "output", paste0("ModelLog_", resp_var, "_", dat, "_")), Sys.Date(), ".txt"))
model
print("Final Model")
paste0("ME = ", validation[1,1])
paste0("RMSE = ", validation[1,2])
paste0("R2 = ", validation[1,3])
paste0("CCC = ", validation[1,4])
paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels")
sink()
```

