---
title: "Creating a raster stack of predictor layers"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here, 
       terra
       )
```


## Create directories

```{r create_dirs}
dir.create(here("1_predictors", "data", "input"), recursive = TRUE)
dir.create(here("1_predictors", "data", "output"), recursive = TRUE)
```


# Data processing

## Create template rasters
Create two template raster files with resolutions of 0.05 deg. x 0.05 deg. (template_raster_low) and 0.025 deg. x 0.025 deg. (template_raster_high).

nrows = (ymax - ymin)/resolution
ncols = (xmax - xmin)/resolution

```{r templates}
template_raster_low <- rast(nrows = 280, ncols = 264, xmin = -9.8, xmax = 3.4, ymin = 48.2, ymax = 62.2, crs = "EPSG:4326", vals = 1)
plot(template_raster_low)
```


## Load predictor layers

*Ultimately, the predictor layers should be stored at Zenodo, downloaded and unzipped before loading them into the project.*
*It might also be neater to load the predictor files in a loop.*
*The individual layers would also require more meaningful names.*

bathymetry        Emodnet_bathymetry_2024_0.01deg
mud               bias_adjusted_mud
sand              bias_adjusted_sand
gravel            bias_adjusted_gravel
curr_vel_max      max_velocity_original_coarse
curr_vel_mean     mean_velocity_original_coarse
temp              bottom_temp_original
sal               bias_corrected_downscaled_salinity
chl               bias_corrected_downscaled_chl
spm_summer        summer_SPM_original
spm_winter        winter_bias_corrected_downscaled_spm

```{r load_predictors}
temp <- rast(here("1_predictors", "data", "input", "bottom_temp_original.tif"))
chl <- rast(here("1_predictors", "data", "input", "bias_corrected_downscaled_chl.tif"))
mud <- rast(here("1_predictors", "data", "input", "bias_adjusted_mud.tif"))
gravel <- rast(here("1_predictors", "data", "input", "bias_adjusted_gravel.tif"))
names(temp) <- "temp"
names(chl) <- "chl"
```


## Create additive log-ratios of sediment composition data

Mud, sand, and gravel content are compositional data that must sum to 100% (or 1 as fractional data). Projecting these data individually will lead to results that do not adhere to this. A workaround is to calculate additive log-ratios, project these and the calculate mud, sand, and gravel content from the projected additive log-ratios.

```{r create_alr}
alr_m <- log(mud/gravel)
alr_s <- log(sand/gravel)
```


## Project predictor layers

*Maybe this could also be looped.*

method	
character. Method used for estimating the new cell values of a SpatRaster. One of:

near: nearest neighbor. This method is fast, and it can be the preferred method if the cell values represent classes. It is not a good choice for continuous values. This is used by default if the first layer of x is categorical.
bilinear: bilinear interpolation. This is the default if the first layer of x is numeric (not categorical).
cubic: cubic interpolation.
cubicspline: cubic spline interpolation.
lanczos: Lanczos windowed sinc resampling.
sum: the weighted sum of all non-NA contributing grid cells.

```{r project_predictors}
temp_pr <- project(x = temp, y = template_raster_low)
chl_pr <- project(x = chl, y = template_raster_low)
```


## Back-transform additive log-ratios

mud = exp(alr_m)/(exp(alr_m) + exp(alr_s) + 1)
sand = exp(alr_s)/(exp(alr_m) + exp(alr_s) + 1)
gravel = 1 - (mud + sand)

```{r back-transform}

```



## Create a raster stack

```{r raster_stack}
env_vars <- c(temp_pr, chl_pr)
plot(env_vars)
```

## Load area of interest

```{r load_aoi}
aoi <- vect(here("1_predictors", "data", "input", "UK_shelf.shp"))

```


## Crop raster stack to UK shelf extent

touches = TRUE?

```{r crop_env_vars}
env_vars <- crop(x = env_vars, y = aoi, mask = TRUE)
plot(env_vars)
```


## Blank out areas mapped as bedrock

The predictive seabed sediments map from the BGS includes bedrock at the seabed surface. Such areas should be blanked out in the raster stack.

```{r bedrock}
#sbs <- vect(here("1_predictors", "data", "input", "Predictive_SBS_UK_V1_ESRI", "Classified_Seabed_Sediments_Folk_UK_v1.shp"))
#rock <- subset(sbs, sbs$Folk_Class == "Rock")


```


# Save raster stack

```{r save_raster_stack}
writeRaster(x = env_vars, filename = here("1_predictors", "data", "output", "env_vars.tif"), overwrite = TRUE)
```



