---
title: "Creating a raster stack of predictor layers with data from BIO-Oracle"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
#devtools::install_github("bio-oracle/biooracler")

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here, 
       terra,
       )
```


# Data processing

## Define resolution and extent of template raster

Resolution and corners of the extent in decimal degrees.

```{r resolution}
r <- 0.05
xmin <- -9.8
xmax <- 3.4
ymin <- 48.2
ymax <- 62.2
```


## Create template rasters

```{r template}
template_raster <- rast(nrows = (ymax - ymin)/r, ncols = (xmax - xmin)/r, xmin = xmin, xmax = xmax, ymin = ymin, ymax = ymax, crs = "EPSG:4326", vals = 1)
```


## Load locally stored predictor layers

```{r load_predictors}

# List all .tif files in the input folder
tif_files <- list.files(here("1_predictors", "data", "input", "BIO-Oracle"), pattern = "\\.tif$", full.names = TRUE)

# Create a vector with variable names based on file names. Will be used when creating the rasterstack.
var_names <- tools::file_path_sans_ext(basename(tif_files))

# Import each .tif file and assign it to a variable named after the file
for (file in tif_files) {
  var_name <- tools::file_path_sans_ext(basename(file))
  assign(var_name, rast(file))
}
```


## Create additive log-ratios of sediment composition data

Mud, sand, and gravel content are compositional data that must sum to 100% (or 1 as fractional data). Projecting these data individually will lead to results that do not adhere to this. A workaround is to calculate additive log-ratios, project these and the calculate mud, sand, and gravel content from the projected additive log-ratios. 

### Replace zeros

However, as log(0) is undefined, we first have to replace any 0 with a small number.

```{r replace_zeros}
MUD_BGS_none[MUD_BGS_none == 0] <- 0.001
SAND_BGS_none[SAND_BGS_none == 0] <- 0.001
GRAV_BGS_none[GRAV_BGS_none == 0] <- 0.001
```


### Rescale

We now need to assure that sums are 100%. We do this by rescaling.

```{r sum_and rescale}
SUM <- MUD_BGS_none + SAND_BGS_none + GRAV_BGS_none
MUD_BGS_none <- MUD_BGS_none * 100 / SUM
SAND_BGS_none <- SAND_BGS_none * 100 / SUM
GRAV_BGS_none <- GRAV_BGS_none * 100 / SUM
SUM2 <- MUD_BGS_none + SAND_BGS_none + GRAV_BGS_none
plot(SUM2)
```


### Create ALRs

```{r create_alr}
alr_m <- log(MUD_BGS_none/GRAV_BGS_none)
alr_s <- log(SAND_BGS_none/GRAV_BGS_none)
```


## Project predictor layers

*Maybe this could also be looped.*

method	
character. Method used for estimating the new cell values of a SpatRaster. One of:

near: nearest neighbor. This method is fast, and it can be the preferred method if the cell values represent classes. It is not a good choice for continuous values. This is used by default if the first layer of x is categorical.
bilinear: bilinear interpolation. This is the default if the first layer of x is numeric (not categorical).
cubic: cubic interpolation.
cubicspline: cubic spline interpolation.
lanczos: Lanczos windowed sinc resampling.
sum: the weighted sum of all non-NA contributing grid cells.

```{r project_predictors}
bath_pr       <- project(x = BATH_EMODnet_none, y = template_raster)
dist_pr       <- project(x = DIST_calc_none, y = template_raster)
alr_m_pr         <- project(x = alr_m, y = template_raster)
alr_s_pr        <- project(x = alr_s, y = template_raster)

bwv_max_pr  <- project(x = Present.Benthic.Mean.Depth.Current.Velocity.Max.tif.BOv2_1, y = template_raster)
bwv_mean_pr <- project(x = Present.Benthic.Mean.Depth.Current.Velocity.Mean.tif.BOv2_1, y = template_raster)
bwv_min_pr <- project(x = Present.Benthic.Mean.Depth.Current.Velocity.Min.tif.BOv2_1, y = template_raster)
bwv_range_pr <- project(x = Present.Benthic.Mean.Depth.Current.Velocity.Range.tif.BOv2_1, y = template_raster)

bwo_max_pr <- project(x = Present.Benthic.Mean.Depth.Dissolved.oxygen.Max, y = template_raster)
bwo_mean_pr <- project(x = Present.Benthic.Mean.Depth.Dissolved.oxygen.Mean, y = template_raster)
bwo_min_pr <- project(x = Present.Benthic.Mean.Depth.Dissolved.oxygen.Min, y = template_raster)
bwo_range_pr <- project(x = Present.Benthic.Mean.Depth.Dissolved.oxygen.Range, y = template_raster)

bws_max_pr <- project(x = Present.Benthic.Mean.Depth.Salinity.Max, y = template_raster)
bws_mean_pr <- project(x = Present.Benthic.Mean.Depth.Salinity.Mean, y = template_raster)
bws_min_pr <- project(x = Present.Benthic.Mean.Depth.Salinity.Min, y = template_raster)
bws_range_pr <- project(x = Present.Benthic.Mean.Depth.Salinity.Range, y = template_raster)

bwt_max_pr <- project(x = Present.Benthic.Mean.Depth.Temperature.Max, y = template_raster)
bwt_mean_pr <- project(x = Present.Benthic.Mean.Depth.Temperature.Mean, y = template_raster)
bwt_min_pr <- project(x = Present.Benthic.Mean.Depth.Temperature.Min, y = template_raster)
bwt_range_pr <- project(x = Present.Benthic.Mean.Depth.Temperature.Range, y = template_raster)

chla_max_pr <- project(x = Present.Surface.Chlorophyll.Max, y = template_raster)
chla_mean_pr <- project(x = Present.Surface.Chlorophyll.Mean, y = template_raster)
chla_min_pr <- project(x = Present.Surface.Chlorophyll.Min, y = template_raster)
chla_range_pr <- project(x = Present.Surface.Chlorophyll.Range, y = template_raster)

pp_max_pr <- project(x = Present.Surface.Primary.productivity.Max, y = template_raster)
pp_mean_pr <- project(x = Present.Surface.Primary.productivity.Mean, y = template_raster)
pp_min_pr <- project(x = Present.Surface.Primary.productivity.Min, y = template_raster)
pp_range_pr <- project(x = Present.Surface.Primary.productivity.Range, y = template_raster)

```


## Back-transform additive log-ratios

The additive log ratios can noow be back-transformed to mud, sand and gravel content.

```{r back-transform}
mud_pr = exp(alr_m_pr)/(exp(alr_m_pr) + exp(alr_s_pr) + 1)
sand_pr = exp(alr_s_pr)/(exp(alr_m_pr) + exp(alr_s_pr) + 1)
gravel_pr = 1 - (mud_pr + sand_pr)

# Test to check that transformation worked
sum_pr <- mud_pr + sand_pr + gravel_pr
plot(sum_pr)
```



## Create a raster stack

```{r raster_stack}
env_vars <- c(
  bath_pr,
  dist_pr,
  mud_pr,
  sand_pr,
  gravel_pr,
  bwv_max_pr,
  bwv_mean_pr,
  bwv_min_pr,
  bwv_range_pr,
  bwo_max_pr,
  bwo_mean_pr,
  bwo_min_pr,
  bwo_range_pr,
  bws_max_pr,
  bws_mean_pr,
  bws_min_pr,
  bws_range_pr,
  bwt_max_pr,
  bwt_mean_pr,
  bwt_min_pr,
  bwt_range_pr,
  chla_max_pr,
  chla_mean_pr,
  chla_min_pr,
  chla_range_pr,
  pp_max_pr,
  pp_mean_pr,
  pp_min_pr,
  pp_range_pr
  )

names(env_vars) <- var_names
```


## Blank out areas mapped as bedrock

The predictive seabed sediments map from the BGS includes bedrock at the seabed surface. Such areas should be blanked out in the raster stack.

```{r bedrock}
sbs <- vect(here("1_predictors", "data", "input", "BIO-Oracle", "Classified_Seabed_Sediments_Folk_UK_v1.shp"))
SBS <- rasterize(sbs, template_raster, field = "Folk_Class")
SBS[SBS$Folk_Class == "Rock"] <- NA

env_vars <- crop(x = env_vars, y = SBS, mask = TRUE)
```


## Load area of interest

```{r load_aoi}
aoi <- vect(here("1_predictors", "data", "input", "BIO-Oracle", "UK_shelf.shp"))
```


## Crop raster stack to UK shelf extent

```{r crop_env_vars}
env_vars <- crop(x = env_vars, y = aoi, mask = TRUE)
plot(env_vars)
```


# Save raster stack

```{r save_raster_stack}
writeRaster(x = env_vars, filename = here("1_predictors", "data", "output", paste0("env_vars_bio-oracle_", r, "deg.tif")), overwrite = TRUE)
```


