---
title: "Organic carbon content - modelling"
output: html_notebook
---

# Install packages

```{r packages, message=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here,
       terra,
       dplyr,
       caret,
       ggplot2,
       sf,
       CAST,
       lwgeom,
       geosphere,
       quantregForest,
       doParallel,
       ModelMetrics,
       DescTools,
       forcats)
```


# Preparation

## Load required data

```{r load_data}
predictors <- rast(here("2_OC_content_model", "data", "interim", "predictors.tif"))

OC <- vect(here("2_OC_content_model", "data", "interim", "OC.shp"))

rm_resp <- read.csv(here("2_OC_content_model", "data", "interim", "rm_resp.csv"))
summary(rm_resp)
```


## Which response variable are we predicting?

Define response variable, unit and data:

all: OCdata_extended.csv
ll: unprojected (latitude and longitude)
0.01: resolution of predictors in degrees

```{r response_type}
resp_var <- "OC"
resp_unit <- "weight-%"
dat <- "all_ll_0.01"
```


# Quantile Regression Forest model

## Distances in environmental space

Distances in environmental (feature) space are computed.

```{r env_space_dist, message=FALSE}
dist_env <- geodist(st_as_sf(OC), 
                    predictors,
                    type = "feature"
                     )

plot(dist_env)
plot(dist_env) + scale_x_log10()
```


## Distances in geographic space

Distances in geographic space are computed.

```{r geogr_space_dist, message=FALSE}
dist_geogr <- geodist(st_as_sf(OC), 
                     predictors,
                     type = "geo"
                     )

plot(dist_geogr)
plot(dist_geogr, unit="km") + scale_x_log10()
```


## Creating spatial blocks

Spatial blocks and folds are created. The folds will be used in a spatial k-fold cross validation. The k-fold nearest neighbour distance matching algorithm is used here.

```{r knndm}
k <- 5 # Number of folds
knndmfolds <- knndm(tpoints = st_as_sf(OC),
                    modeldomain = predictors,
                    k = k,
                    samplesize = 2000)
```


## Distances in geographic space including CV distances

```{r geogr_space_dist2, message=FALSE}
dist_geogr2 <- geodist(st_as_sf(OC), 
                       predictors,
                       cvfolds= knndmfolds$indx_test,
                       type = "geo"
                       )

plot(dist_geogr2)
plot(dist_geogr2, unit="km") + scale_x_log10()
```


## Model tuning

A Quantile Regression Forest model is tuned. Predictor variables are selected in a forward feature selection approach and various values of the mtry parameter are tested in a spatial k-fold cross validation.

The maximum number of iterations to be performed can be calculated upfront, based on the number of pre-selected predictors:

```{r max_iter}
factorial(length(names(predictors)))/(factorial(2)*factorial(length(names(predictors))-2)) + sum(c((length(names(predictors))-2):1))
```


### Forward feature selection

The best combination of predictor variables (features) is found in a forward feature selection process.

```{r ffs, message=FALSE, warning=FALSE}
nCores <- detectCores()
cl <- makePSOCKcluster(nCores - 1)
registerDoParallel(cl)

set.seed(42)

model <- ffs(rm_resp[names(predictors)],
               rm_resp$OC_pc,
               metric = "Rsquared",
               method="qrf",
               what = 0.5,
               replace = FALSE,
               importance = TRUE,
               trControl = trainControl(method="CV",
                                        number = k,
                                        savePredictions = "final",
                                        index = knndmfolds$indx_train, 
                                        allowParallel = TRUE),
               verbose = TRUE)

stopCluster(cl)

model

sel_preds <- model$selectedvars
```


### FFS plot

Plot of R2 over the model runs.

```{r ffs_plot}
plot(model)
```


## Validation statistics

The validation results of the optimal RF model.

Note that these are the statistics based on the predicted values of the selected model. These differ from the values from the tuning (above), which are the means of the k predictions based on the folds.

```{r validation_stats}
t <- data.frame(model$pred$pred, model$pred$obs)

ccc <- CCC(x = t$model.pred.obs, y = t$model.pred.pred) # Concordance Correlation Coefficient

validation <- data.frame(mse=numeric(), rmse=numeric(), r2=numeric())
validation[1,1] <- round(sum(t$model.pred.obs - t$model.pred.pred)/nrow(t), 3)
validation[1,2] <- round(rmse(t$model.pred.obs, t$model.pred.pred), 3)
validation[1,3] <- round(cor(t$model.pred.obs, t$model.pred.pred)^2, 3)
validation[1,4] <- round(ccc$rho.c$est, 3)

colnames(validation) <- c("ME", "RMSE", "r2", "CCC")
rownames(validation) <- NULL
validation
```


## Histograms of observed and predicted organic carbon content (weight-%)

```{r histogram_oc, message=FALSE}
hist <- ggplot(t) + 
  geom_histogram(aes(x = model.pred.obs), fill = "#69b3a2", alpha = 0.5) +
  geom_label( aes(x=5, y=1500, label="observed"), color="#69b3a2") +
  geom_histogram(aes(x = model.pred.pred), fill = "#404080", alpha = 0.5) +
  geom_label( aes(x=5, y=1350, label="predicted"), color="#404080") +
  theme_bw() +
  xlab(paste0(resp_var, " (", resp_unit, ")")) +
  ylab("count")
  
hist

jpeg(filename = here("2_OC_content_model", "figures", paste0(resp_var, "_obs_vs_pred_histogram.jpg")), width = 15, height = 10, units = "cm", res = 300)
hist
dev.off()
```


## Validation plot

```{r validation_plot, message=FALSE}
p <-  ggplot(t, aes(x = model.pred.pred, y = model.pred.obs)) +
      geom_hex(bins = 50) +
      #geom_point() +
      scale_fill_continuous(type = "viridis") +
      geom_abline(intercept = 0, slope = 1, colour = "grey", linewidth = 1) +
      geom_smooth(method = "lm") +
      theme_bw() +
      scale_x_continuous(name = "Predicted value") +
      scale_y_continuous(name = "Observed value") +
      ggtitle(paste0(resp_var, " (", resp_unit, ")"))

p

jpeg(filename = here("2_OC_content_model", "figures", paste0("validation_plot_",resp_var, ".jpg")), width = 15, height = 10, units = "cm", res = 300)
p
dev.off()
```


## Variable importance

```{r variable_importance_plot, warning=FALSE}
imp <- varImp(model$finalModel, scale = FALSE)
if (!is.data.frame(imp) && "importance" %in% names(imp)) imp <- imp$importance
imp <- as.data.frame(imp)
imp$Predictor <- rownames(imp)
rownames(imp) <- NULL

importance_col <- if ("Overall" %in% names(imp)) "Overall" else names(imp)[1]
imp <- imp[order(imp[[importance_col]], decreasing = TRUE), c("Predictor", importance_col)]
colnames(imp)[2] <- "IncMSE"

impfig <- imp %>%
  mutate(Predictor = fct_reorder(Predictor, IncMSE)) %>%
  ggplot( aes(x=Predictor, y=IncMSE)) +
    geom_bar(stat="identity", fill="#f68060", alpha=.6, width=.4) +
    coord_flip() +
    xlab("") +
    ylab("% increase in MSE") +
    theme_bw()
    
impfig

jpeg(filename = here("2_OC_content_model", "figures", paste0("variable_importance_plot_",resp_var, ".jpg")), width = 10, height = 10, units = "cm", res = 300)
impfig
dev.off()
```


## Distances in environmental space including CV distances

```{r env_space_dist2, message=FALSE}
dist_env2 <- geodist(st_as_sf(OC), 
                     predictors,
                     type = "feature",
                     cvfolds= knndmfolds$indx_test,
                     variables = sel_preds
                     )

plot(dist_env2)
plot(dist_env2) + scale_x_log10()
```


## Partial dependence

Partial dependence plots give a graphical depiction of the marginal effect of a variable on the response.

```{r partial_plots}
m2 <- model$finalModel
class(m2) <- "randomForest"

for (i in 1:length(sel_preds)) {
  partialPlot(x = m2, pred.data = rm_resp, x.var = sel_preds[i], main = "", xlab = sel_preds[i], ylab = paste0(resp_var, " (", resp_unit, ")"))
}
```


# Predict QRF model

## Predict respons

The response variable is predicted. Median values of the QRF distribution are calculated as central values. Different percentiles are predicted as well.

```{r predict_oc}
preds <- raster::stack(predictors[[sel_preds]])
med <- predict(preds, model$finalModel, what = 0.5)
p5 <- predict(preds, model$finalModel, what = 0.05)
p25 <- predict(preds, model$finalModel, what = 0.25)
p75 <- predict(preds, model$finalModel, what = 0.75)
p95 <- predict(preds, model$finalModel, what = 0.95)

```


## Area of applicability

```{r aoa}
trainDI <- trainDI(model = model,
                        variables = sel_preds)
print(trainDI)

aoa <- aoa(newdata = predictors, 
                model = model,
                trainDI = trainDI,
                variables = sel_preds,
)

plot(aoa)

fr <- freq(aoa$AOA)
print(paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels"))
```


## Plot results

```{r plot_results}
plot(rast(p5), main = paste0(resp_var, " P5"))
plot(rast(p25), main = paste0(resp_var, " P25"))
plot(rast(med), main = paste0(resp_var, " median"))
plot(rast(p75), main = paste0(resp_var, " P75"))
plot(rast(p95), main = paste0(resp_var, " P95"))
plot(aoa$DI, main = "Dissimilarity index")
plot(aoa$AOA, main = "Area of applicability")
# Uncertainty maps (QRF prediction interval width)
PI_abs <- mask(rast(p95) - rast(p5), aoa$AOA, maskvalue = 0)
PI_rel <- mask((rast(p95) - rast(p5)) / rast(med) * 100, aoa$AOA, maskvalue = 0)
plot(PI_abs, main = paste0(resp_var, " uncertainty abs (P95-P5)"))
plot(PI_rel, main = paste0(resp_var, " uncertainty rel (P95-P5)/median [%]"))
```


## Convert AOA from raster to polygon

```{r aoa_poly}
aoa_poly <- as.polygons(aoa$AOA, dissolve = TRUE)
plot(aoa_poly)

write_sf(st_as_sf(aoa_poly), dsn = here("2_OC_content_model", "data", "output"), layer = paste0(resp_var, "_", dat, "_AOA_", Sys.Date()), driver = "ESRI Shapefile")
```


## Export results

```{r export_results}
writeRaster(med, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_median_", Sys.Date(), ".tif"))
writeRaster(p5, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_P5_", Sys.Date(), ".tif"))
writeRaster(p25, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_P25_", Sys.Date(), ".tif"))
writeRaster(p75, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_P75_", Sys.Date(), ".tif"))
writeRaster(p95, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_P95_", Sys.Date(), ".tif"))
writeRaster(aoa$DI, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_DI_", Sys.Date(), ".tif"))
writeRaster(aoa$AOA, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_AOA_", Sys.Date(), ".tif"))
writeRaster(PI_abs, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_PI_abs_", Sys.Date(), ".tif"))
writeRaster(PI_rel, paste0(here("2_OC_content_model", "data", "output"), "/", resp_var, "_", dat, "_PI_rel_", Sys.Date(), ".tif"))
```


## Output a log file

```{r log}
sink(file = paste0(here("2_OC_content_model", "data", "output", paste0("ModelLog_", resp_var, "_", dat, "_")), Sys.Date(), ".txt"))
model
print("Final Model")
paste0("ME = ", validation[1,1])
paste0("RMSE = ", validation[1,2])
paste0("R2 = ", validation[1,3])
paste0("CCC = ", validation[1,4])
paste0("AOA = ", round(100*fr$count[2]/ sum(fr$count),2), "% of pixels")
sink()
```

