---
title: "Organic carbon content - data exploration"
output: html_notebook
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here, 
       CAST,
       corrplot,
       ggplot2)
```


## Load required data

```{r load_data}
rm_resp <- read.csv(here("2_OC_content_model", "data", "interim", "rm_resp.csv"))

predictors <- rast(here("1_predictors", "data", "output", "env_vars_bio-oracle_0.05deg.tif"))
#predictors <- predictors[[1:11]] # Remove winter SPM
```


# Data exploration

## Histogram of organic carbon content (weight-%)

```{r histogram_oc_stock, message=FALSE}
hist <- ggplot(rm_resp) + 
  geom_histogram(aes(x = OC_pc), fill = "#69b3a2", alpha = 0.8) +
  theme_bw() +
  xlab("OC content (weight-%)") +
  ylab("count")

hist

jpeg(filename = here("2_OC_content_model", "figures", "OC_obs_histogram.jpg"), width = 15, height = 10, units = "cm", res = 300)
hist
dev.off()
```


## Correlation plot

```{r corr_plot}
corrplot.mixed(cor(rm_resp[, names(predictors)]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)

jpeg(filename = here("2_OC_content_model", "figures", "correlation_plot.jpg"), width = 30, height = 30, units = "cm", res = 300)
corrplot.mixed(cor(rm_resp[, names(predictors)]), lower.col =  "black", tl.pos = "lt", number.cex = 0.6)
dev.off()
```


## Environmental space

A visual check to what extent the samples cover the environmental space. This is useful as legacy data were used and no formal sampling design was applied in the analysis.

* Blue: Samples

* Grey: Environmental data (based on random subsample)

```{r env_space_sf}
smp <- as.data.frame(spatSample(x = predictors, size = nrow((rm_resp)), method = "random", na.rm = TRUE))


for (i in names(predictors)) {
    
  print(ggplot() +
          geom_density(data = rm_resp, aes(x=rm_resp[,i]),colour="cornflowerblue",fill="cornflowerblue", alpha=0.1,linewidth=1) +
          geom_density(data = smp, aes(x=smp[,i]), colour="grey",fill="grey", alpha=0.1, linewidth=1) +
          scale_x_continuous(name = names(rm_resp[i])))
        
}
```




