---
title: "Dry bulk density - data preparation"
output:
  html_document:
    df_print: paged
---

# Preparations

## Install packages

```{r packages, message=FALSE, warning=FALSE}
rm(list=ls())

if(!require("pacman")) install.packages("pacman"); library(pacman)
p_load(here,
       sf,
       spatialEco,
       dplyr,
       stars,
       terra)
```


## Define projection and resolution

```{r projection}
crs <- "EPSG:4326" # WGS84
res <- 0.01
```


## Which response variable are we predicting?

```{r response_type}
resp_var <- "DBD"
```


# Predictor variables

Potentially relevant predictor variables are loaded.

```{r load_predictors}
predictors <- rast(here("1_predictors", "data", "output", "env_vars_0.01deg.tif"))
predictors <- predictors[[1:11]] # Remove winter SPM
#OC <- rast(here("2_OC_content_model", "data", "output", "OC_all_ll_0.01_median_2025-12-18.tif"))
#predictors <- c(predictors, OC)
#names(predictors)[dim(predictors)[3]] <- "OC"
plot(predictors)

```


## Ensure uniform projection

Check if predictors have the defined projection. Re-project if this is not the case.

```{r uniform_proj}
if (crs(predictors) != crs) {
  predictors <- project(predictors, crs, res = res)
}
```


## Create a fishnet based on predictor raster

A spatial polygons file is created. It consists of one square for every grid cell. It will be used later in the process to create centroids which hold the mean value of the measurements.

```{r fishnet}
min_ext <- sum(predictors)
min_ext[min_ext == 0] <- NA
fishnet <- st_as_sf(stars::st_as_stars(min_ext), as_points = FALSE, merge = FALSE)
fishnet$ID <- 1:nrow(fishnet)
```


# Response variable

## Load response

The response data is divided in two data sets from two different organizations. Here we show the code to import both and merge them into one data set. The data set is then converted to a spatial object.    

1. St Andrews data set:
```{r load_response}
resp_data_1 <- read.csv(here("3_DBD_model", "data", "input", "dbd.csv"), header = TRUE, sep = ",")
resp_data_1 <- resp_data_1[,c(2,3,4,7)]
summary(resp_data_1)

# Convert to sf
resp_1 <- st_as_sf(resp_data_1, coords = c("longitude", "latitude"), crs = "EPSG:4326")

summary(resp_1)
```
2. CEFAS data set:
```{r load_response}
resp_data_2 <- read.csv(here("3_DBD_model", "data", "input", "Poseidon_BC_externalCollaboration.csv"), header = TRUE, sep = ",")
resp_data_2 <- resp_data_2[,c(1,2,3,9)]
summary(resp_data_2)

# Convert to sf
resp_2 <- st_as_sf(resp_data_2, coords = c("X", "Y"), crs = "EPSG:4326")

summary(resp_2)
```

Now we will bind these two data sets together to start the modelling process
```{r}
head(resp_1)
names(resp_1) <- names(resp_1)
head(resp_2)
names(resp_2) <- names(resp_1)
resp <- rbind(resp_1, resp_2, deparse.level =1)
```

## Ensure uniform projection

Check if the response variable has the defined projection. Re-project if this is not the case.

```{r uniform_proj_CRI}
if (st_crs(resp)$proj4string != crs) {
  resp <- st_transform(resp, crs)
}
```


## Centroids

Centroids are calculated to average all measurements within one grid cell.

```{r centroids_surface, warning=FALSE}
centroids <- st_join(fishnet, resp)
centroids <- na.omit(centroids)
centroids <- centroids %>% group_by(ID) %>% summarize(DBD = mean(dbd_g_cm.3, na.rm = TRUE))
resp <- st_centroid(centroids)

plot(predictors$BATH_EMODnet_none)
plot(resp[,1], pch = 20, col = "black", cex = 0.5, add = TRUE)
```


## Create a regression matrix

```{r regression_matrix}
rm_resp <- as.data.frame(terra::extract(predictors, resp, bind = TRUE))
rm_resp <- rm_resp[-1] #removes first column (ID)
rm_resp <- na.omit(rm_resp) #removes rows with NAs
summary(rm_resp)
```


# Save files for further use

```{r save_files}
write_sf(resp, dsn = here("3_DBD_model", "data", "interim"), layer = resp_var, driver = "ESRI Shapefile")
write.csv(rm_resp, here("3_DBD_model", "data", "interim", "rm_resp.csv"), quote = FALSE, row.names = FALSE)
writeRaster(predictors, here("3_DBD_model", "data", "interim", "predictors.tif"), overwrite = TRUE)
```

